---
title: "EMPAFunctionAnalysis Demo"
author: "Duy Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EMPAFunctionAnalysis Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
devtools::load_all(".")
library(EMPAFunctionAnalysis)
library(dplyr)
```

This vignette demonstrates the core scoring functions of the `EMPAFunctionAnalysis` package. All functions accept raw data directly — no pre-cleaning required.

---

# Load Data

```{r load-data}
vegetativecover_data <- read.csv(
  "https://nexus.sccwrp.org/empachecker/export?tablename=tbl_vegetativecover_data"
)
vegetation_sample_metadata <- read.csv(
  "https://nexus.sccwrp.org/empachecker/export?tablename=tbl_vegetation_sample_metadata"
)
```

```{r}
cram <- read.csv("inst/extdata/EMPA_CRAM_2014-2024.csv")
gis_data <- read.csv("inst/extdata/BufferLandCover_251203.csv")
wetland_extents <- read.csv("inst/extdata/HabitatZones_260224.csv")
ruggedness <- read.csv("inst/extdata/Ruggedness_250702.csv")
```

---

# Score All Metrics at Once

The `score_all_metrics()` function runs every metric and returns a single long-format table. Use the `year` argument to filter to specific years.

```{r score-all}
all_scores <- score_all_metrics(
  vegetation_sample_metadata = vegetation_sample_metadata,
  vegetativecover_data = vegetativecover_data,
  cram = cram,
  gis_data = gis_data,
  wetland = wetland_extents,
  rugged = ruggedness,
  year = c(2023, 2024)
)

all_scores
```

```{r score-all-table}
all_scores |>
  count(function_name, indicator_name, metric_name) |>
  arrange(function_name, indicator_name, metric_name)
```

---

# Individual Metrics

## Plant Function

### CRAM Index

Normalized CRAM score per site. Left-joined onto vegetation sites so every surveyed site appears — `NA` where no CRAM data exists.

```{r cram-index}
cram_scores <- score_cram_index(
  cram = cram,
  vegetativecover_data = vegetativecover_data,
  function_name = "Plant",
  year = c(2023)
)

head(cram_scores)
```

### Native Cover

Percent of total cover that is native, after excluding unknown species and non-recorded statuses.

```{r native-cover}
native_scores <- score_native_cover(
  vegetativecover_data = vegetativecover_data,
  year = c(2023)
)

head(native_scores)
```

### Invasive Severity

Starts at 100 and subtracts penalty points per unique Cal-IPC-rated invasive species. Floored at 0.

```{r invasive-severity}
invasive_scores <- score_invasive_severity(
  vegetativecover_data = vegetativecover_data,
  year = c(2023)
)

head(invasive_scores)
```

### Vegetated Cover

Percent of total cover that is vegetated, summed across all plots per site.

```{r veg-cover}
veg_cover_scores <- score_veg_cover(
  vegetation_sample_metadata = vegetation_sample_metadata,
  function_name = "Plant",
  year = c(2023)
)

head(veg_cover_scores)
```

### Ruggedness

Pre-computed surface ruggedness index passed through directly. Static — no year variation.

```{r ruggedness}
rugged_scores <- score_ruggedness(ruggedness)

head(rugged_scores)
```

---

## SLR Function

### Buffer Cover (500 m)

Percent natural landcover (Ag + Natural) within the 500 m buffer around each site.

```{r buffer-cover}
buffer_scores <- score_buffer_cover(
  gis_data = gis_data,
  buffer_size = "500 m",
  metric_name = "buffer_cover"
)

head(buffer_scores)
```

### Perimeter Land Cover (30 m)

Same calculation as buffer cover, using the 30 m buffer to capture the immediate site perimeter.

```{r perimeter-land-cover}
perimeter_scores <- score_buffer_cover(
  gis_data = gis_data,
  buffer_size = "30 m",
  metric_name = "perimeter_land_cover"
)

head(perimeter_scores)
```

### Perimeter Contiguity

Ratio of the largest contiguous open patch to total open area. Higher = less fragmented.

```{r perimeter-contiguity}
contiguity_scores <- score_perimeter_contiguity(gis_data)

head(contiguity_scores)
```

### Current Habitat Distribution

Percent of the current wetland footprint covered by vegetated marsh (Low + Mid + High zones).

```{r current-extent}
current_scores <- score_current_habitat_distribution(wetland_extents)

head(current_scores)
```

### Future Habitat Distribution

Same calculation on the modeled 1.2 ft sea level rise migration footprint (avoiding developed land).

```{r future-extent}
future_scores <- score_future_habitat_distribution(wetland_extents)

head(future_scores)
```

---

# Filtering by Year

All year-varying metrics accept a `year` argument. Pass a vector to include multiple years, or `"all"` to include everything.

```{r multi-year}
# Multiple years
native_multi <- score_native_cover(
  vegetativecover_data = vegetativecover_data,
  year = c(2021, 2023)
)

native_multi |>
  count(year)
```

```{r all-years}
# All years (default)
native_all <- score_native_cover(
  vegetativecover_data = vegetativecover_data
)

native_all |>
  count(year) |>
  arrange(year)
```
